Supplement: Bootstrapping Domain Specific Languages for Bayesian Program Learning

### 2018-02-09 supplement.tex
no more errors
edits to supplement

\usepackage{listings}
\usepackage{amsthm}
% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfig} 
\usepackage{fancyvrb}

@@

\usepackage{verbatim}
 \usepackage{booktabs}
 % For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tikz}
\usetikzlibrary{fit,bayesnet}
%\usetikzlibrary{arrows.meta}
%\usetikzlibrary{positioning}
%\usetikzlibrary{decorations.text}
%\usetikzlibrary{decorations.pathmorphing}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{hyperref}
\DeclareMathOperator*{\argmin}{arg\,min} % thin space, limits underneath in displays
\DeclareMathOperator*{\argmax}{arg\,max} % thin space, limits underneath in displays
\DeclareMathOperator{\argmin}{argmin} % no space, limits underneath in displays

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \begin{scope}[shift={(2.5,0)}]
    \node[obs] at (3,3) (dx){$\mathcal{D}$};
    \node[latent] at (3.5,1.75) (zp){$p$};
    \node[obs] at (4,3) (tx){$\theta^{(x)}$};
    \node[obs] at (3.5,0.7) (xp) {$x$};
    \node[align = center] at ([yshift = -0.6cm,xshift = 0.5cm]xp.south) {Wake: Infer $p$};
    \draw [->,red] (xp.east) to[out = 0,in = 0] node(nn){} (tx.east);
    \draw [->,red] (tx) -- (zp);
    \draw [->] (dx) -- (zp);
    \node at (nn) {
      \begin{tikzpicture}[x=2.5cm,y=1.25cm,transform canvas={scale=0.2,shift={+(-1,2.5)}}]
        \tikzstyle{neuron}=[circle,fill=blue!50,minimum size=20pt]
        \fill[fill=white] (-0.25,-0.5) rectangle (2.25,-4.5);
        \node[rectangle,width=2cm] at (1,1) {};
        \foreach \name / \y in {1,...,4}
            \node[neuron] (I-\name) at (0,-\y) {};
        \foreach \name / \y in {1,...,3}
            \node[neuron] (H-\name) at (1,-\y-0.5) {};
        \foreach \name / \y in {1,...,4}
            \node[neuron] (O-\name) at (2,-\y) {};
        \foreach \source in {1,...,4}
            \foreach \dest in {1,...,3}
                \draw [-latex] (I-\source) -- (H-\dest);
        \foreach \source in {1,...,3}
            \foreach \dest in {1,...,4}
                \draw [-latex] (H-\source) -- (O-\dest);
      \end{tikzpicture}
    };
    \node[shift={+(0,-0.65)}] at (nn) [fill=white]{ $q(x)$ };
  \end{scope}

@@

  \begin{scope}[shift={(7.7,-4)}]
    \node[latent] at (0.5,3) (d){$\mathcal{D}$};
    \node[latent] at (1.5,3) (t){$\theta$};
    \node[obs] at (1,1.75) (z){$p$};
    % \node[latent] at (3.5,1.5) (tx){$\theta^{(x)}$};
    \node[obs] at (1,0.5) (x) {$x$};
    \edge {z}{x};
    \edge {d,t}{z};
    \plate {}{(z)(x)}{$x\in X$};
    \node[align = center] at ([yshift = -1cm]x.south) {Sleep-G: Induce $(\mathcal{D},\theta})$};
  \end{scope}

@@

\section{Program Representation}\label{programrepresentation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\\\noindent A \emph{primitive}, like the number 5 or
  the function \texttt{sum}.
\\\noindent A \emph{variable}, like $x$, $y$, $z$
\\\noindent A $\lambda$\emph{-abstraction}, which creates a new function. $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$.
\\\noindent An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$.

For example, the function which squares the logarithm of a number is
$\lambda x.\text{\texttt{square}} (\text{\texttt{log} } x)$, and the identity function $f(x) = x$ is $\lambda x.x$. The
$\lambda$-calculus serves as a spartan but expressive Turing complete
program representation, and distills the essential features of functional languages like Lisp.

However, many $\lambda$-calculus expressions correspond to ill-typed programs, such as the program that takes the logarithm of the Boolean \texttt{true} (i.e., \texttt{log true}) or which applies the number five to the identity function
(i.e., $5 \; (\lambda x.x)$).
We use a well-established typing system for $\lambda$-calculus called \emph{Hindley-Milner typing}~\cite{pierce}, which is used in programming languages like OCaml.
The purpose of the typing system is to ensure that our programs never call a function with a type it is not expecting (like trying to take the logarithm of \texttt{true}).
Hindley-Milner has two important features:
Feature 1: It supports \emph{parametric polymorphism}: meaning that types can have variables in them, called \emph{type variables}. Lowercase Greek letters are conventionally used for  type variables.
For example, the type of the identity function is $\alpha\to\alpha$, meaning it takes something of type $\alpha$ and return something of type $\alpha$. A function that returns the first element of a list has the type $\texttt{list}(\alpha)\to\alpha$. Type variables are not the same has variables introduced by $\lambda$-abstractions.
Feature 2: Remarkably, there is a  simple algorithm for automatically inferring the polymorphic Hindley-Milner type of a $\lambda$-calculus expression~\cite{damas1982principal}.
Our generative model over programs performs Hindley-Milner type inference during sampling;
\texttt{unify} in the generative model uses the machinery of Hindley-Milner to
ensure that the generated programs have valid polymorphic types.
A satisfactory exposition of Hindley-Milner is beyond the scope of this paper,
but~\cite{pierce} offers a nice overview of lambda calculus and typing systems like Hindley-Milner.

@@

We use Algorithm~\ref{grammarInductionAlgorithm} to search for the structure of the DSL that best explains the frontiers.

\begin{algorithm}[tb]
   \caption{DSL Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - ||\theta||_0$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ subexpression of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \STATE \textbf{if }$\text{score}(\mathcal{D}') < \text{score}(\mathcal{D})$\textbf{ return }$\mathcal{D},\theta^*(\mathcal{D})$
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}

@@

We write $c(e,p)$ to mean the number of times that primitive $e$ was used in program $p$;  $R(p)$ to mean the sequence of types input to sample in Alg.~1 of the main paper. Jensen's inequality gives an intuitive lower bound on the likelihood of a program $p$:
\begin{align*}
  \log \probability[p|\theta]&\stackrel{+}{ = }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - \sum_{\tau\in R(p)} \log \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  &\stackrel{+}{\geq }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{\tau\in R(p)} \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  & = \sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{e\in \mathcal{D}} r(e,p)\theta_e
\end{align*}
where $c(p) = \sum_{e\in \mathcal{D}}c(e,p)$ and $r(e:\tau',p) = \sum_{\tau\in R(p)} \indicator[\text{canUnify}(\tau,\tau')]$.

Differentiate with respect to $\theta_e$ and set to zero 
\begin{align}
  \frac{c(x)}{\theta_x} &= N\frac{a(x)}{\sum_y a(y)\theta_y}
\end{align}
This equality holds if $\theta_x = c(x)/a(x)$:
\begin{align}
  \frac{c(x)}{\theta_x} &= a(x).\\
N\frac{a(x)}{\sum_y a(y)\theta_y}& = N\frac{a(x)}{\sum_y c(y)}   = N\frac{a(x)}{N} = a(x).
\end{align}
If this equality holds then $\theta_x \propto c(x)/a(x)$:
\begin{align}
  \theta_x = \frac{c(x)}{a(x)}\times \underbrace{\frac{\sum_y a(y)\theta_y}{N}}_{\text{Independent of $x$}}.
\end{align}

Now what we are actually after is the parameters that maximize the joint log probability of the data and parameters $\theta$, which we  write $J$:
\begin{align}
  J& = L + \log \text{D}(\theta|\alpha)\\
  &\stackrel{+}{\geq } \sum_x c(x)\log \theta_x - N \log \sum_x a(x)\theta_x  + \sum_x(\alpha_x - 1)\log \theta_x\\
  & = \sum_x (c(x) + \alpha_x - 1)\log \theta_x -  N \log \sum_x a(x)\theta_x
\end{align}
We identify the parameters of the prior over $\theta$, $\alpha$, as  pseudocounts that are added to the \emph{counts} ($c(x)$), but not to the \emph{possible counts} ($a(x)$).

When is this lower bound tight? This lower bound is tight whenever all
of the types of the expressions in the DSL are not polymorphic, in which case our DSL is equivalent to a PCFG
and this estimator is equivalent to the inside/outside algorithm.
Polymorphism introduces context-sensitivity to the DSL,
and exactly maximizing the likelihood with respect to $\theta$
becomes intractable,
so for domains with polymorphic types we use this estimator.

@@

If we were to maximize the ELBO bound to perform inference in
our generative model,
then, during DSL induction, we would seek a new $(\mathcal{D}^*,\theta^*)$ maximizing the following lower bound on the likelihood (along an unimportant regularizing term on the DSL):
\begin{align}
  &  \sum_{x\in X}\expect_{p\sim Q_x}\left[\log \probability[p|\mathcal{D}^*,\theta^*] \right] \label{EM}\\
  & Q_x(p)\triangleq \probability[p|x,\mathcal{D},\theta]
\end{align}
where $(\mathcal{D},\theta)$ is our current estimate of the generative
model.  These equations fall out of an EM-style derivation, and one
could replace $Q_x(p)$ with the recognition model $q(p|x)$, either
using importance sampling (so the expectation in~\ref{EM} is taken
over $q$ and we reweight using $Q_x$) or by directly using $q$ as our approximate posterior over the program that solves task $x$.

How the recognition model is worked into this is unimportant to the reason why we do not maximize a bound of this form:
this bound takes an expectation over the \emph{previous} sleep cycle's
posterior over programs,
and so the approximate posterior $Q_x$ at the next iteration
ends up being very close to previous approximate posterior.
Intuitively,
we want the DSL induction to be a function \emph{only} of the programs that we have found,
and \emph{not} be a function of how the previous generative model weighted them.
In practice,
we found that maximizing EM-style bounds, like the ELBO,
leads to a kind of hysteresis effect,
where the next generative model too closely matches the previous one,
causing the algorithm to easily become trapped in local optima.

>>>>

\usepackage{listings}
\usepackage{amsthm}
% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
%\usepackage{subfig}
\usepackage{fancyvrb}

@@

\usepackage{verbatim}
 \usepackage{booktabs}
 % For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tikz}
\usetikzlibrary{fit,bayesnet}
%\usetikzlibrary{arrows.meta}
%\usetikzlibrary{positioning}
%\usetikzlibrary{decorations.text}
%\usetikzlibrary{decorations.pathmorphing}
\usepackage{dsfont}
\usepackage{amsmath}
\usepackage{hyperref}
%\DeclareMathOperator*{\argmin}{arg\,min} % thin space, limits underneath in displays
\DeclareMathOperator*{\argmax}{arg\,max} % thin space, limits underneath in displays
\DeclareMathOperator{\argmin}{argmin} % no space, limits underneath in displays

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \begin{scope}[shift={(2.5,0)}]
    \node[obs] at (3,3) (dx){$\mathcal{D}$};
    \node[latent] at (3.5,1.75) (zp){$p$};
    \node[obs] at (4,3) (tx){$\theta^{(x)}$};
    \node[obs] at (3.5,0.7) (xp) {$x$};
    \node[align = center] at ([yshift = -0.6cm,xshift = 0.5cm]xp.south) {Wake: Infer $p$};
    \draw [->,red] (xp.east) to[out = 0,in = 0] node(nn){} (tx.east);
    \draw [->,red] (tx) -- (zp);
    \draw [->] (dx) -- (zp);
    \node at (nn) {
      \begin{tikzpicture}[x=2.5cm,y=1.25cm,transform canvas={scale=0.2,shift={+(-1,2.5)}}]
        \tikzstyle{neuron}=[circle,fill=blue!50,minimum size=20pt]
        \fill[fill=white] (-0.25,-0.5) rectangle (2.25,-4.5);
        \node[rectangle] at (1,1) {};
        \foreach \name / \y in {1,...,4}
            \node[neuron] (I-\name) at (0,-\y) {};
        \foreach \name / \y in {1,...,3}
            \node[neuron] (H-\name) at (1,-\y-0.5) {};
        \foreach \name / \y in {1,...,4}
            \node[neuron] (O-\name) at (2,-\y) {};
        \foreach \source in {1,...,4}
            \foreach \dest in {1,...,3}
                \draw [-latex] (I-\source) -- (H-\dest);
        \foreach \source in {1,...,3}
            \foreach \dest in {1,...,4}
                \draw [-latex] (H-\source) -- (O-\dest);
      \end{tikzpicture}
    };
    \node[shift={+(0,-0.65)}] at (nn) [fill=white]{ $q(x)$ };
  \end{scope}

@@

  \begin{scope}[shift={(7.7,-4)}]
    \node[latent] at (0.5,3) (d){$\mathcal{D}$};
    \node[latent] at (1.5,3) (t){$\theta$};
    \node[obs] at (1,1.75) (z){$p$};
    % \node[latent] at (3.5,1.5) (tx){$\theta^{(x)}$};
    \node[obs] at (1,0.5) (x) {$x$};
    \edge {z}{x};
    \edge {d,t}{z};
    \plate {}{(z)(x)}{$x\in X$};
    \node[align = center] at ([yshift = -1cm]x.south) {Sleep-G: Induce $(\mathcal{D},\theta)$};
  \end{scope}

@@

\section{Program Representation}\label{programrepresentation}
We choose to represent programs using $\lambda$-calculus~\cite{pierce}.
A $\lambda$-calculus expression is either:
\begin{itemize}
  \item[--] A \emph{primitive}, like the number 5 or the function \texttt{sum}.
  \item[--] A \emph{variable}, like $x$, $y$, or $z$.
  \item[--] A $\lambda$\emph{-abstraction}, which creates a new function.  $\lambda$-abstractions have a variable and a body. The body is a $\lambda$-calculus expression. Abstractions are written as $\lambda \text{var}. \text{body}$ or in Lisp syntax as \mbox{\texttt{(lambda (\textrm{var}) \textrm{body})}}.
  \item[--] An \emph{application} of a function to an argument. Both the function and the argument are $\lambda$-calculus expressions. The application of the function $f$ to the argument $x$ is written as $f\; x$ or as \texttt{($f$ x)}.
\end{itemize}

For example, the function which squares the logarithm of a number is
$\lambda x.\text{\tt(square (log $x$))}$, and the identity function $f(x) = x$ is $\lambda x.x$. The
$\lambda$-calculus serves as a spartan but expressive Turing complete
program representation, and distills the essential features of functional
programming languages like Lisp.

However, many $\lambda$-calculus expressions correspond to ill-typed programs, such as the program that takes the logarithm of the Boolean \texttt{true} (i.e., \texttt{log true}) or which applies the number five to the identity function
(i.e., $5 \; (\lambda x.x)$).
We use a well-established typing system for $\lambda$-calculus called \emph{Hindley-Milner typing}~\cite{pierce}, which is used in programming languages like OCaml.
The purpose of the typing system is to ensure that our programs never call a function with a type it is not expecting (like trying to take the logarithm of \texttt{true}).
Hindley-Milner has two important features:
Feature 1: It supports \emph{parametric polymorphism}, meaning that types can have variables in them, called \emph{type variables}. Lowercase Greek letters are conventionally used for  type variables.
For example, the type of the identity function is $\alpha\to\alpha$, meaning it takes something of type $\alpha$ and return something of type $\alpha$. A function that returns the first element of a list has the type $[\alpha]\to\alpha$. Type variables are not the same as variables introduced by $\lambda$-abstractions.
Feature 2: Remarkably, there is a  simple algorithm for automatically inferring the polymorphic Hindley-Milner type of a $\lambda$-calculus expression~\cite{damas1982principal}.
Our generative model over programs performs Hindley-Milner type inference during sampling:
\emph{Unify} in the generative model uses the machinery of Hindley-Milner to
ensure that the generated programs have valid polymorphic types.
A satisfactory exposition of Hindley-Milner is beyond the scope of this paper,
but~\cite{pierce} offers a nice overview of lambda calculus and typing systems like Hindley-Milner.

@@

We use Alg.~\ref{grammarInductionAlgorithm} to search for the structure of the DSL that best explains the frontiers.

\setcounter{algorithm}{2} % there are two algorithms in the main paper.
\begin{algorithm}[tb]
   \caption{DSL Induction Algorithm}
   \label{grammarInductionAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
     \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
     \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
     \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
     \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - \|\theta\|_0$
     \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
     \WHILE {true}
     \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ subexpression of }z\}$
     \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
     \STATE \textbf{if }$\text{score}(\mathcal{D}') < \text{score}(\mathcal{D})$\textbf{ return }$\mathcal{D},\theta^*(\mathcal{D})$
     \STATE $\mathcal{D}\gets\mathcal{D}'$
     \ENDWHILE
   \end{algorithmic}
\end{algorithm}

@@

We write $c(e,p)$ to mean the number of times that primitive $e$ was used in program $p$; $c(p)= \sum_{e\in \mathcal{D}}c(e,p)$ to mean the total number of primitives used in program $p$; $R(p)$ to mean the sequence of types input to sample in Alg.~1. Jensen's inequality gives an intuitive lower bound on the likelihood of a program $p$:
\begin{align*}
  \log \probability[p|\theta]&\stackrel{+}{ = }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - \sum_{\tau\in R(p)} \log \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  &\stackrel{+}{\geq }\sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{\tau\in R(p)} \sum_{\substack{e:\tau'\in \mathcal{D}\\\text{unify}(\tau,\tau')}}\theta_e\\
  & = \sum_{e\in \mathcal{D}} c(e,p)\log \theta_e - c(p)\log \sum_{e\in \mathcal{D}} r(e,p)\theta_e
\end{align*}
where  $r(e:\tau',p) = \sum_{\tau\in R(p)} \indicator[\text{canUnify}(\tau,\tau')]$.

Differentiate with respect to $\theta_e$ and set to zero:
%\begin{align*}
%  \frac{c(e, p)}{\theta_e} &= c(p)\frac{r(e, p)}{\sum_{e'\in\mathcal{D}} r(e', p)\theta_{e'}}
%\end{align*}
%or more succinctly,
\begin{align*}
  \frac{c(x,p)}{\theta_x} &= N\frac{a(x)}{\sum_y a(y)\theta_y}
\end{align}
where $N=c(p)$ and $a(x) = r(x, p)$. From here forward, we similarly let
$c(x)=c(x,p)$ for notational convenience.

This equality holds if $\theta_x = c(x)/a(x)$:
\begin{align}
  \frac{c(x)}{\theta_x} &= a(x).\\
N\frac{a(x)}{\sum_y a(y)\theta_y}& = N\frac{a(x)}{\sum_y c(y)}   = N\frac{a(x)}{N} = a(x).
\end{align}
If this equality holds then $\theta_x \propto c(x)/a(x)$:
\begin{align}
  \theta_x = \frac{c(x)}{a(x)}\times \underbrace{\frac{\sum_y a(y)\theta_y}{N}}_{\text{Independent of $x$}}.
\end{align}

Now what we are actually after is the parameters that maximize the joint log probability of the data and parameters $\theta$, which we  write $J$:
\begin{align}
  J& = L + \log \text{Dir}(\theta|\alpha)\\
  &\stackrel{+}{\geq } \sum_x c(x)\log \theta_x - N \log \sum_x a(x)\theta_x  + \sum_x(\alpha_x - 1)\log \theta_x\\
  & = \sum_x (c(x) + \alpha_x - 1)\log \theta_x -  N \log \sum_x a(x)\theta_x
\end{align}
We identify the parameters of the prior over $\theta$, $\alpha$, as  pseudocounts that are added to the \emph{counts} ($c(x)$), but not to the \emph{possible counts} ($a(x)$).

This lower bound is tight whenever all
of the types of the expressions in the DSL are not polymorphic, in which case our DSL is equivalent to a PCFG
and this estimator is equivalent to the inside/outside algorithm.
Polymorphism introduces context-sensitivity to the DSL,
and exactly maximizing the likelihood with respect to $\theta$
becomes intractable,
so for domains with polymorphic types we use this estimator.

@@

If we were to maximize the ELBO bound to perform inference in
our generative model,
then, during DSL induction, we would seek a new $(\mathcal{D}^*,\theta^*)$ maximizing the following lower bound on the likelihood (along an unimportant regularizing term on the DSL):
\begin{align}
  &  \sum_{x\in X}\expect_{p\sim Q_x}\left[\log \probability[p|\mathcal{D}^*,\theta^*] \right] \label{EM}\\
  & Q_x(p)\triangleq \probability[p|x,\mathcal{D},\theta]
\end{align}
where $(\mathcal{D},\theta)$ is our current estimate of the generative
model.  These equations fall out of an EM-style derivation, and one
could replace $Q_x(p)$ with the recognition model $q(p|x)$, either
using importance sampling (so the expectation in Eq.~\ref{EM} is taken
over $q$ and we reweigh using $Q_x$) or by directly using $q$ as our approximate posterior over the program that solves task $x$.

How the recognition model is worked into this is unimportant to the reason why we do not maximize a bound of this form:
this bound takes an expectation over the \emph{previous} sleep cycle's
posterior over programs,
and so the approximate posterior $Q_x$ at the next iteration
ends up being very close to previous approximate posterior.
Intuitively,
we want the DSL induction to be a function \emph{only} of the programs that we have found,
and \emph{not} be a function of how the previous generative model weighed them.
In practice,
we found that maximizing EM-style bounds, like the ELBO,
leads to a kind of hysteresis effect,
where the next generative model too closely matches the previous one,
causing the algorithm to easily become trapped in local optima.
